$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-16 paper/paper.tex
Preliminary pass to the example

% Fix autoref's example treatment 
\newtheorem{example}{Example}
\newcommand{\exampleautorefname}{Example}
\newcommand{\nitheoremautorefname}{Theorem}
\newcommand{\nidefinitionautorefname}{Definition}

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
\renewcommand{\nidefinitionautorefname}{Definition}%
\renewcommand{\nitheoremautorefname}{Theorem}%
\newcommand{\nilemmaautorefname}{Lemma}%
\renewcommand{\sectionautorefname}{\S}%
\renewcommand{\subsectionautorefname}{\S}%

@@

Linearity also influenced the $\pi$-calculus.
Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
Meanwhile, linearity acquired a different flavour with the rise of \emph{session types} \cite{H93,THK94,HVK98}, a formalism used in communication-centric programming, where linearity ensures that a channel is owned by exactly one communicating participant --- the channel itself may be used multiple times, in particular by following its session type.
Linearity in session types is the key ingredient that guarantees communication safety and session fidelity.

@@

\begin{example}[The courier system]
\label{example-process}

@@

A courier system consists of three roles: a sender, who wants to send a package; a receiver, who receives the package sent by the sender and a courier, who carries the package from the sender to the receiver.
Our courier system consists of four \picalc{} processes composed in parallel:
two sender processes, $S_1$ and $S_2$, sending data over channels $x$ and $y$, respectively;
one receiver process, $R$, which receives the data from each of the senders;
and finally a courier process $C$, which synchronises communication among the previous three process and
first receives data from $S_1$ and $S_2$ on its input channels, respectively $x$ and $y$ and then sends the received data to $R$ along its output channel $z$.
A graphical representation of the courier system is given in \autoref{fig:example-process}.

@@

\begin{figure}[t]
  \small
  \centering
  \begin{tikzpicture}[draw,circle]
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-x) at (0,2)   {$\func{send} \, x$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-y) at (2,2)   {$\func{send} \, y$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=green] (recv-z) at (4,2)   {$\func{recv} \, z$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=blue,align=center] (sync) at (2,0)   {$\func{cour}$\\$ \, x \, y \, z$};

@@

The receiver process $R$ is defined as follows:
\begin{flalign*}
  & \func{recv} \; : \; \Var_n \to \Process_n & \\
  & \func{recv} \; c \; = \recv{c}{} (\recv{\suc c} \PO) \\
\end{flalign*}

The sender processes sender processes $S_1$ and $S_2$ are defined as follows:
\begin{flalign*}
  & \func{send} \; : \; \Var_n \to \Process_n & \\
  & \func{send} \; c = \new{} \send{\suc c}{\constr{0}} \PO \\
\end{flalign*}

The courier process $C$ is defined as follows:
\begin{flalign*}
  & \func{cour} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
  & \func{cour} \; in0 \; in1 \; out \; &&= \\
  & && \recv{in0}{} \\
  & && \recv{(\suc \; in1)}{} \\
  & && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
  & && \send{(\suc \suc \; out)}{\constr{0}} \PO \\
\end{flalign*}
\begin{flalign*}
  & \func{courier} && : \; \Process_{\constr{0}} & \\
  & \func{courier} && = \new{} ( \func{send} \; \constr{0} \\
  & && \comp{}{ \new{} ( \func{send} \; \constr{0} } \\
  & && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
  & && \comp{}{ \func{cour} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}
[ALIGN FIG AND PROCESSES IN PARTICULAR send c and send x,y in THE FIGURE]
[COMMENT ON THE PROCESSES AND WHAT THEY DO]
[EXPLAIN HERE WHAT IS TYPED WITH LIN, GRADED AND SHARED]
Thanks to linear and graded types, the above processes can have their interactions safely restricted: both $\func{send}$ processes consume $0$ input and $1$ output multiplicities; $\func{recv}$ consumes $2$ input and $0$ output multiplicities.
Dually, the process $\func{cour}$ consumes one input multiplicity on the shared channels with $\func{send}$, respectively $x$ and $y$ in \autoref{fig:example-process} and two output multiplicities on the channel shared with $\func{recv}$, channel $z$ in \autoref{fig:example-process}.
We use these types to give a suitable typing derivation in \autoref{example-derivation} (\autoref{leftover-typing}).
\end{example}

@@

\begin{example}[Typing derivation]
\label{example-derivation}

@@

\begin{flalign*}
& \func{\vdash sync} && : \types{\gamma \comma \channel{t}{\lz} \comma \channel{t}{\lz} \comma \channel{t}{\lz}\\ & &&}{\Gamma \comma (\suc lx \comma rx) \comma (\suc ly \comma ry) \comma (lz \comma \suc \suc rz)\\ & &&}{\func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}\\ & &&}{\Gamma \comma (lx \comma rx) \comma (ly \comma ry) \comma (lz \comma rz)} & \\
& \func{\vdash sync} && = \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \send{\suc \suc \constr{0}}{\suc \constr{0}} \\
& && \send{\suc \suc \constr{0}}{\constr{0}} \PO \\
\end{flalign*}

@@

\begin{flalign*}
& \func{\vdash example} && : \; \types{\constr{[]}}{\constr{[]}}{\func{example}}{\constr{[]}} & \\
& \func{\vdash example} && = \new{} \; \_ \; \_ \; \_ \; ( \func{\vdash send} \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; ( \func{\vdash send} \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; ( \func{\vdash recv} \\
& && \comp{}{ \func{\vdash sync} }))) \\
\end{flalign*}

>>>>

% Continue examples later on
\usepackage{thmtools}
\newtheorem{example}{Example}
\renewcommand\thmcontinues[1]{Continued}

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
\renewcommand{\nidefinitionautorefname}{Definition}%
\renewcommand{\nitheoremautorefname}{Theorem}%
\renewcommand{\nilemmaautorefname}{Lemma}%
\renewcommand{\sectionautorefname}{\S}%
\renewcommand{\subsectionautorefname}{\S}%
\renewcommand{\exampleautorefname}{Example}

@@

Linearity also influenced the $\pi$-calculus.
Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
Meanwhile, linearity acquired a different flavour with the rise of \emph{session types} \cite{H93,THK94,HVK98}, a formalism used in communication-centric programming, where linearity ensures that a channel is owned by exactly one communicating participant --- the channel itself may be used multiple times, as specified by its session type.
Linearity in session types is the key ingredient that guarantees communication safety and session fidelity.

@@

\begin{example}[The courier system, label=exa:cont]
\label{example-process}

@@

A courier system consists of three roles: a sender, who wants to send a package; a receiver, who receives the package sent by the sender; and a courier, who carries the package from the sender to the receiver.
Our courier system consists of four \picalc{} processes composed in parallel:
two sender processes, $S_1$ and $S_2$, sending data over channels $x$ and $y$, respectively;
one receiver process, $R$, which receives the data from each of the senders;
and a courier process $C$, which synchronises communication among the previous three process.
The courier process $C$ first receives data from $S_1$ and $S_2$ on its input channels $x$ and $y$, respectively, and then sends the received data to $R$ along its output channel $z$.
A graphical representation of the courier system is given in \autoref{fig:example-process}.

@@

\begin{figure}[t]
  \small
  \centering
  \begin{tikzpicture}[draw,circle]
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-x) at (0,2)   {$\func{send} \, x$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-y) at (2,2)   {$\func{send} \, y$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=green] (recv-z) at (4,2)   {$\func{recv} \, z$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=blue,align=center] (sync) at (2,0)   {$\func{carry}$\\$ \, x \, y \, z$};

@@

The receiver process $R$, defined below, receives data twice on a channel $c$ and then terminates.
\begin{flalign*}
  & \func{recv} \; : \; \Var_n \to \Process_n & \\
  & \func{recv} \; c \; = \recv{c}{} (\recv{\suc c} \PO) \\
\end{flalign*}

The sender processes $S_1$ and $S_2$, defined below, create a new channel that acts as the data that is being sent, send it over the channel $c$, and then terminate.
\begin{flalign*}
  & \func{send} \; : \; \Var_n \to \Process_n & \\
  & \func{send} \; c = \new{} \send{\suc c}{\constr{0}} \PO \\
\end{flalign*}

The courier process $C$, defined below, sequentially receives on the two input channels $in0$ and $in1$, and then outputs the two pieces of received data on the output channel $out$.
\begin{flalign*}
  & \func{carry} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
  & \func{carry} \; in0 \; in1 \; out \; &&= \\
  & && \recv{in0}{} \\
  & && \recv{(\suc \; in1)}{} \\
  & && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
  & && \send{(\suc \suc \; out)}{\constr{0}} \PO \\
\end{flalign*}

Finally, we fit these processes together and create three communication channels: the first shared between the sender $S_1$ and the courier $C$, the second between the sender $S_2$ and the courier $C$, and the third between the receiver $R$ and the courier $C$.
The result is the process defined below.
\begin{flalign*}
  & \func{system} && : \; \Process_{\constr{0}} & \\
  & \func{system} && = \new{} ( \func{send} \; \constr{0} \\
  & && \comp{}{ \new{} ( \func{send} \; \constr{0} } \\
  & && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
  & && \comp{}{ \func{carry} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}

[ALIGN FIG AND PROCESSES IN PARTICULAR send c and send x,y in THE FIGURE]

We continue this example by providing typing derivations for these processes in \autoref{leftover-typing}.
We use graded types for the usage annotations of the communication channels $x$, $y$, and $z$: both processes $\func{send} \; x$ and $\func{send} \; y$ consume $1$ output multiplicity; $\func{recv} \; z$ consumes $2$ input multiplicities.
Dually, the process $\func{carry} \; x \; y \; z$ consumes one input multiplicity on $x$ and $y$, and two output multiplicities on $z$.

[TODO: type of data being transmitted]
\end{example}

@@

\begin{example}[Typing derivation, continues=exa:cont]
\label{example-derivation}

@@

\begin{flalign*}
& \func{\vdash carry} && : \types{\gamma \comma \channel{t}{\lz} \comma \channel{t}{\lz} \comma \channel{t}{\lz}\\ & &&}{\Gamma \comma (\suc lx \comma rx) \comma (\suc ly \comma ry) \comma (lz \comma \suc \suc rz)\\ & &&}{\func{carry} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}\\ & &&}{\Gamma \comma (lx \comma rx) \comma (ly \comma ry) \comma (lz \comma rz)} & \\
& \func{\vdash carry} && = \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \send{\suc \suc \constr{0}}{\suc \constr{0}} \\
& && \send{\suc \suc \constr{0}}{\constr{0}} \PO \\
\end{flalign*}

@@

\begin{flalign*}
& \func{\vdash system} && : \; \types{\constr{[]}}{\constr{[]}}{\func{system}}{\constr{[]}} & \\
& \func{\vdash system} && = \new{} \; \_ \; \_ \; \_ \; ( \func{\vdash send} \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; ( \func{\vdash send} \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; ( \func{\vdash recv} \\
& && \comp{}{ \func{\vdash carry} }))) \\
\end{flalign*}
